---
#output:
#  html_document:
#    keep_md: true
#output: 
#  pdf_document:
#    keep_tex: true
#output: word_document
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path="Figs/", warning=FALSE, message=FALSE)
```


# Stuff You Should Know About: Handling Large Data Files with R
## The Data Table package and various other ways to handle data in R

 - **Authors**: Tom Kelly
- **Research field**: Bioinformatics / Computational Biology / Cancer Genomics
 - **Lesson Topic**: An introduction to various packages for file I/O and data manipulation in R, with comparision to base R (and compatibility with data frames), in terms of user-friendliness, performance in CPU-time, and memory usage. 

## Installation

Install Data Table from CRAN (current version 1.9.6)
```{r}
install.packages("data.table", repos = "https::/cran.rstudio.com")
library("data.table")
```

Install development version from GitHub (current version 1.9.7)
```{r, eval=FALSE}
install.packages("data.table", repos = "https://Rdatatable.github.io/data.table", type = "source") #v1.9.7
library("data.table")
```

## Getting Started: Data Frames

data table has it's own read function - to rapidly read data into R
Backwards compatible: It can be used for data.frames
```{r}
gapminderFiveYearData <- fread("gapminder-FiveYearData.csv", data.table=F)
class(gapminderFiveYearData)
dim(gapminderFiveYearData)
head(gapminderFiveYearData)
tail(gapminderFiveYearData)
str(gapminderFiveYearData)
```

Backwards compatible: these are standard dataframes compatible with ggplots
```{r}
library("ggplot2")
ggplot(data = gapminderFiveYearData, aes(x = lifeExp, y = gdpPercap, color=continent)) +
  geom_point()
```

##Introducing Data Tables

data table defaults to reading it's own data.table format
```{r}
gapminderFiveYearData <- fread("gapminder-FiveYearData.csv")
class(gapminderFiveYearData)
dim(gapminderFiveYearData)
head(gapminderFiveYearData)
tail(gapminderFiveYearData)
str(gapminderFiveYearData)
```
Data tables also auto-trim when printing to console
```{r}
gapminderFiveYearData
```

data tables are backwards compatible with a lot of operations which use data.frames
Such as plots...
```{r}

ggplot(data = gapminderFiveYearData, aes(x = lifeExp, y = gdpPercap, color=continent)) +
  geom_point()
```
... and linear models...
```{r}
linear_model <- lm(gdpPercap ~ pop + year, gapminderFiveYearData)
summary(linear_model)
linear_model <- lm(lifeExp ~ gdpPercap + pop + year, gapminderFiveYearData)
summary(linear_model)
linear_model <- glm(lifeExp ~ gdpPercap + continent + pop + year, family  ="gaussian", gapminderFiveYearData)
summary(linear_model)
```
... and data manipulation packages (plyr, dplyr, reshape, tidyr, etc...)
```{r}
library("plyr")
calcGDP <- function(dat, year=NULL, country=NULL) {
  if(!is.null(year)) {
    dat <- dat[dat$year %in% year, ]
  }
  if (!is.null(country)) {
    dat <- dat[dat$country %in% country,]
  }
  gdp <- dat$pop * dat$gdpPercap
  
  new <- cbind(dat, gdp=gdp)
  return(new)
}
plyr::ddply(
  .data = calcGDP(gapminderFiveYearData),
  .variables = "continent",
  .fun = function(x) mean(x$gdp)
)
```
Yeah you get the idea.

Data tables have built-in "methods" for a range of functions, these are often faster than standard dataframes or matrices, if these aren't found it uses dataframe functions. A "Data Table" is compatible with any function from any package designed for a "Data Frame".

```{r, include=FALSE}
#generate gapminder-large data
source("gapminderlarger.R")
fwrite(gapminderlarge, file="gapminder-large.csv")
fwrite(gapminderlarger, file="gapminder-larger.csv")
```


## File I/O (Input/Output)
fread is "fast read", and it's **fast**, even for large data files. Let's try it out on some larger datafiles:
```{r}
gapminderlarge <- fread("gapminder-large.csv", header=T)
```
fread is smart, it auto detects column classes, separators, headers, nrows (for a regularly separated file). We can use the same comand for a whole bunch of file formats. All the usual reading options can be specified manually...
```{r}
gapminderFiveYearData <- fread("gapminder-FiveYearData.tsv") #tab delimited
gapminderFiveYearData <- fread("gapminder-FiveYearData.txt") #space delimited
gapminderFiveYearDataCrop <- fread("gapminder-FiveYearData.tsv", header=T, col.names=c("place", "time", "people", "big place", "life", "money"), nrows=1000, stringsAsFactors=F)
gapminderFiveYearDataCrop
```
...but it does a lot of the tedious work for you (pretty well too).

It's also got cool progress bars for large files :) These kick in automatically if the file takes longer than about a second. This is really handy to know your code is working, and how long it will take.
```{r}
gapminderlarger <- fread("gapminder-larger.csv")
```
It's so fast it tells you. Let's compare that with base R:
```{r}
system.time(gapminderlarger.dataframe <- read.csv("gapminder-larger.csv", header=T))
```
The same operation took much longer with base R, with larger files (or repeating this many times) that ~6x difference could mean a lot for your workflow. 

FYI - there's also a "fast write" compatible with several file formats
```{r}
fwrite(gapminderlarger, file="test.csv") #defaults to csv
fwrite(gapminderlarger, file="test.tsv", sep="\t")
```
They're also fast to write data, compared to base R:
```{r}
system.time(fwrite(gapminderlarger, file="test.csv"))
system.time(write.csv(gapminderlarger, file="test.csv"))
```

## readr (Hadley Wickham and RStudio)
Another package enables faster alternatives to existing read functions in base R: these work almost exactly the same as their base R counterparts.

 | **base R** | **readr**
--- | --- | ---
spaced file | `read.table`   | `read_table`
fixed-width file | `read.fwf`   | `read_fwf`
comma-separated file | `read.csv`   | `read_csv`
semicolon-separated file | `read.csv2`   | `read_csv2`
tab-delimited file | `read.table`   | `read_tsv`
comma-separated file | `read.csv`   | `read_csv`
file or string `readLines`   | `read_lines` or `read_file`

Let's try it out on a space-delimited file:
```{r}
library("readr")
system.time(read_table("gapminder-FiveYearData.txt"))
system.time(read.table("gapminder-FiveYearData.txt"))
```
Even on a small file `readr` is faster than base R. This also holds for larger csv files:
```{R}
system.time(read_csv("gapminder-larger.csv"))
system.time(read.csv("gapminder-larger.csv"))
```
`readr` also has a handy progress bar allowign us to monitor progress. There is an equivalent `readxl` package with a `read_excel` function compatible with xls or xlsx files and enables sheet selection. This is a relatively new alternative to the `xlsx` package and it's `read.xlsx` function which are difficult to work with (as it is java and perl dependent).

## Another solution: bigmemory
```{r}
library("bigmemory")
```
"bigmemory" uses the "big.matrix" format to access large data files in a C++ framework - rather than stored in RAM/memory as usual in R. This is handy for handling **very large** files, when loading the full dataset in working environment (RAM memory) slows your computer to a halt. Might be handy on servers / HPC too but usually they have enough memory if you're willing to wait for it in a queue.

Let's try out bigmemory, first we convert an R data matrix into a "big.matrix":
```{r}
gapminderFiveYearData.big <- as.big.matrix(gapminderFiveYearData)
gapminderFiveYearData.big
class(gapminderFiveYearData.big)
dim(gapminderFiveYearData.big)
head(gapminderFiveYearData.big)
tail(gapminderFiveYearData.big)
str(gapminderFiveYearData.big)
```
bigmemory, also has read/write functions direct to big.matrix format:
```{r}
write.big.matrix(gapminderFiveYearData.big, "test.csv")
gapminderFiveYearData.big <- read.big.matrix("test.csv")
```
These are designed to be efficient for memory - how fast are they?
```{r}
system.time(gapminderlarger.big <- read.big.matrix("gapminder-larger.csv"))
system.time(write.big.matrix(gapminderFiveYearData.big, "test.csv"))
```

## New and Shiny: FEATHER
### A Fast On-Disk Format for Data Frames for R and Python, powered by Apache Arrow

FEATHER (is it's own fast file format) - from Hadley Wickham ggplot/dplyr/etc... and Wes Mckinney (pandas in Python)
Note: it's in development (unstable) - future versions may not read past versions - intended for use to transfer files quickly (e.g., between R and Python)

At the moment you can only try it out from their github repo (in R or python), it will no doubt end up on CRAN very soon:
```{r}
library("devtools")
devtools::install_github("wesm/feather/R")
library(feather)
```

FEATHER has it's own file I/O commands (and format):
```{r}
path <- "gapminder-FiveYearData.feather"
write_feather(gapminderFiveYearData, path) #write data frame to file
gapminderFiveYearData <- read_feather(path) #read to data frame
gapminderFiveYearData
```
Did I mention it's crazy fast?
```{r}
path <- "gapminderlarger.feather"
system.time(write_feather(gapminderlarger, path))
system.time(gapminderlarger.feather <- read_feather(path))
```

Or install and run in Python:
```
import feather
path = 'my_data.feather'
feather.write_dataframe(df, path)
df = feather.read_dataframe(path)
```

Note that FEATHER is designed for data _already_ loaded into python or R.

## FILE I/O Summary

### READ 
**base R** | **data table** | **readr** | **bigmemory** | **feather**
--- | --- | --- | --- | ---
`read.csv`   | `fread`         | `read_csv` | `read.big.matrix` | `read_feather`
52.203s | 8.154s | 11.120s | 28.647s | 2.414s

### Convert dataframe to format 
**base R** | **data table** | **bigmemory** | **feather**
--- | --- | --- | ---
`data.frame` | `as.data.table` | `as.big.matrix`   | built-in
NA | 0.002s | 66.07s | NA

### Write 
**base R** | **data table** | **bigmemory** | **feather**
--- | --- | --- | ---
`write.csv`  | `fwrite`       | `write.big.matrix` | `write_feather`
71.382s | 35.453s | 0.068ss | 5.008s

##Manipulating Data Tables
```{r}
gapminderFiveYearData <- fread("gapminder-FiveYearData.csv", data.table=T, header = T)
class(gapminderFiveYearData)
```
We can simply treat it as a data frame in many cases:
```{r}
gapminderFiveYearData[1,]
colnames(gapminderFiveYearData)
head(gapminderFiveYearData$country)
tail(gapminderFiveYearData$country)
```
Data Table has a "Natural" Syntax

`DT[where, select|update|do, by]`

...although suspiciously similar to SQL?

it allows chaining queries: `DT[][]`

Formally: we subset a datatable, Dt, with `DT[i, j, by]`

###I: row selection
```{r}
gapminderFiveYearData[c(1:5, 100:105),] #by number
gapminderFiveYearData[gapminderFiveYearData$country=="New Zealand",] #by condition
gapminderFiveYearData[gapminderFiveYearData$country %in% c("New Zealand", "Australia", "Japan"),] #by condition
gapminderFiveYearData[year=="1952"]
setkey(gapminderFiveYearData, country)
gapminderFiveYearData[c("New Zealand","Australia")] #by key (will be detailed later)
```
###J: column selection
```{r}
head(gapminderFiveYearData[,"country"]) #by names
head(gapminderFiveYearData[,country]) #by column
gapminderFiveYearData[,list(country, year, pop)] #by list
```
This allows operations to be performed on columns:
```{r}
gapminderFiveYearData[,sum(gdpPercap)] #by colnames
gapminderFiveYearData[,sum(gdpPercap*pop)] #by colnames
gapminderFiveYearData[,mean(pop)] #by colnames
gapminderFiveYearData[,mean(lifeExp)] #by colnames
```
###BY: group operation
This is paricularly power in that we can apply operations to sets values, grouped "by":
```{r}
gapminderFiveYearData[j=sum(gdpPercap), by=year]
gapminderFiveYearData[,sum(gdpPercap), year]
gapminderFiveYearData[,mean(lifeExp), year]
gapminderFiveYearData[,sum(pop), by=list(continent, year)]
```
As you can see, these results lend well to data we can tabulate or plot:
```{r}
library("gplots")
plot(gapminderFiveYearData[,sum(pop), by=list(continent, year)]$year,
     gapminderFiveYearData[,sum(pop), by=list(continent, year)]$V1,
     col=rainbow(5)[as.numeric(as.factor(gapminderFiveYearData[,sum(pop), by=list(continent, year)]$continent))])
legend("topleft", fill=rainbow(5), legend=levels(as.factor(gapminderFiveYearData[,sum(pop), by=list(continent, year)]$continent)))
```
New and Shiny: by=.EACHI enables more explicit control of the "by" feature. We could manually pull out years or countries we wish to deal with individually:
```{r}
gapminderFiveYearData[year=="1952" | year=="2002", j=sum(pop), by=year]
gapminderFiveYearData[c("New Zealand","Australia"),sum(gdpPercap*pop)]
gapminderFiveYearData[c("New Zealand","Australia"),sum(gdpPercap*pop), by=year]
```
Notice in both of the above cases the countries are grouped together. Unless specified countries will not be grouped, we can do this either explicitly `by=country` or use the `.EACHI` options for more complex `i` queries:
```{r}
gapminderFiveYearData[c("New Zealand","Australia"),sum(gdpPercap*pop), by=country]
gapminderFiveYearData[c("New Zealand","Australia"),sum(gdpPercap*pop), by=.EACHI]
```
Group by multiple arguments explicitly may also give data in a more sensible format:
```{r}
gapminderFiveYearData[c("New Zealand","Australia"),sum(gdpPercap*pop), by=list(year, country)]
```
`by=.EACHI` is a little weird, it's an explicit way of restoring a previous version `data.table` functionality. Consider a simple operation of counting the rows returned:

By default data.table counts all rows returned: 
```{r}
gapminderFiveYearData[c("New Zealand","Australia"), .N]
```
To restore previous functionality (an implicit by), `.by=.EACHI` will count the number of rows returned _for each_ i. Basically data.table was really clever and did it for you but some people took issue with a by being performed when it wasn't specified.
```{r}
gapminderFiveYearData[c("New Zealand","Australia"), .N, by=.EACHI]
```

## Keys
`tables()` shows all tables and their SQL-like "keys", by default to keys are given:
```{r}
gapminderFiveYearData <- fread("gapminder-FiveYearData.csv")
tables()
```
We can create a unique identifier as a key:
```{r}
rowID <- paste(gapminderFiveYearData$country, gapminderFiveYearData$year)
head(rowID)
tail(head(rowID))
gapminderFiveYearData$rowID <- rowID
gapminderFiveYearData
setkey(gapminderFiveYearData, rowID)
tables()
```
We can search rows `i` for this key:
```{r}
gapminderFiveYearData["New Zealand 1952",] #search row by key
```
In contrast to dataframes (rownames) duplicate keys are permitted:
```{r}
setkey(gapminderFiveYearData, country)
gapminderFiveYearData["New Zealand",]
```
By default, alls rows are returned for each group (rather than only first for dataframe), the `mult="first"` or `"last"` can modify this:
```{r}
gapminderFiveYearData["New Zealand", mult="first"] 
gapminderFiveYearData["New Zealand", mult="last"] 
```

Queries in data.tables aren't just *easier* they're **faster**
```{r}
gapminderFiveYearData["New Zealand", mult="first"] 
system.time(gapminderFiveYearData["New Zealand", mult="first"]) #time 0.001s
gapminderFiveYearData.dataframe <- as.data.frame(gapminderFiveYearData)
gapminderFiveYearData.dataframe[gapminderFiveYearData.dataframe$country=="New Zealand",][1,]
system.time(gapminderFiveYearData.dataframe[gapminderFiveYearData.dataframe$country=="New Zealand",][1,])
```
Ok, that didn't seem that different. They're powerful with larger datafiles though. Compare these examples for the same operation with dataframes and datatables.
```{r}
setkey(gapminderlarger, country)
gapminderlarger["New Zealand", mult="first"] 
system.time(gapminderlarger["New Zealand", mult="first"])
gapminderlarger.dataframe <- as.data.frame(gapminderlarger)
gapminderlarger.dataframe[gapminderlarger.dataframe$country=="New Zealand",][1,]
system.time(gapminderlarger.dataframe[gapminderlarger.dataframe$country=="New Zealand",][1,])
```
Here's an example with multiple keys:
```{r}
setkey(gapminderlarger, country, year)
gapminderlarger[list("New Zealand", 2007)]
system.time(gapminderlarger[list("New Zealand", 2007)])
head(gapminderlarger.dataframe[gapminderlarger.dataframe$country=="New Zealand" & gapminderlarger.dataframe$year=="2007",])
system.time(gapminderlarger.dataframe[gapminderlarger.dataframe$country=="New Zealand" & gapminderlarger.dataframe$year=="2007",])
```

`by` is faster than a simliar operation on dataframes too:
```{r}
gapminderlarger[,sum(gdpPercap), year]
system.time(gapminderlarger[,sum(gdpPercap), year])
tapply(gapminderlarger.dataframe$gdpPercap,gapminderlarger.dataframe$year,sum)
system.time(tapply(gapminderlarger.dataframe$gdpPercap,gapminderlarger.dataframe$year,sum))
```

```{r, include=FALSE}
#export md to README
system("cat README.short syska-R-data-table.md > README.md ")
system("cat README.short syska-R-data-table.tex > README.tex")
```
